Automating Mendelian randomization through machine learning to construct a putative causal map of the human phenome

### 2017-12-11 docs/manuscript.rmd

Mendelian randomization (MR) [@DaveySmith2003; @DaveySmithHemani2014] exploits vertical pleiotropy to infer the causal relationships between phenotypes. Suppose that one trait (the exposure) causally influences another (the outcome). If a SNP influences the outcome through the exposure then the SNP is exhibiting vertical pleiotropy (see Box 1). Such a genetic variant is considered to be a valid instrumental variable if it only influences the outcome through the exposure (the exclusion restriction assumption). Vertical pleiotropy can be exploited to mimic a randomised controlled trial, enabling a causal estimate to be made by comparing the outcome phenotypes between those individuals that have the exposure-increasing allele against those who do not, although not without caveats [@Holmes2017]. Multiple independent genetic variants for a particular exposure can be used jointly to improve causal inference, because a) each variant represents an independent natural experiment, and an overall causal estimate can be obtained by meta-analysing the single estimates from each instrument; and b) potential bias arising from the exclusion restriction assumption can be detected or corrected by evaluating the consistency of effects across instruments [@Bowden2015; @Bowden2016b; @Bowden2017; @Hartwig2017; @Hemani2017; @Verbanck2017].

@@

Problems with obtaining unbiased causal effects can arise, however, if the genetic instruments exhibit horizontal pleiotropy, where they influence the outcome through a pathway other than the exposure (Box 1). The extent of this phenomenon is not to be understated, and many methods have been developed that attempt to reliably obtain unbiased causal estimates under specific models of horizontal pleiotropy [@Bowden2015; @Bowden2016b; @Bowden2017; @Hartwig2017; @Verbanck2017; @Zhu2017]. It is considered best practice to triangulate estimates from a range of MR methods (and other experimental designs) when presenting causal estimates so as to assess the potential for different horizontal pleiotropy models and other violations of MR assumptions [@Lawlor2017]. However there are circumstances in which it is desirable to consolidate across many methods to obtain a single MR estimate. First, it could be of interest to screen links between thousands of traits for being potential causal associations, in which case a critical evaluation of each causal estimate of interest may not be possible or convenient. Second, though the simplest method, based on inverse-variance weighted (IVW) meta analysis [@Pierce2013], is most statistically powerful under no horizontal pleiotropy, it can have high false positive or low true positive rates in the presence of horizontal pleiotropy compared to other methods. Pleiotropy has been hypothesised to be universal [@Wright1968], though the degree and nature (e.g. horizontal versus vertical, see Box 1) to which this may be the case is contested [@Wagner2011; @Hill2012a], hence defaulting to the IVW method in the first instance and using other methods as sensitivity analyses may not be appropriate. Third, if different methods disagree it is not possible to know which is correct because the true nature of horizontal pleiotropy exhibited by the instruments is not known. Fourth, the available methods do not cover all possible models of horizontal pleiotropy, and therefore an automated method for instrument selection may be necessary.

@@

**Mode-based methods:** The mode-based estimator clusters the instruments into groups based on similarity of causal effects, and returns the final causal effect estimate based on the cluster that has the largest number of instruments [@Hartwig2017]. There are four implementations of this method: the simple and the weighted mode, each weighted with or without the assumption of no measurement error in the exposure estimates (NOME). The simple mode is the unweighted mode of the empirical density function of causal estimates, whereas the weighted mode is weighted by the inverse variance of the outcome effect. The bandwidth parameter was set to 1 by default.

@@

We also performed a look up of associations that causally influence years of schooling [@Okbay2016], or that years of schooling itself causally influences. Using a false-discovery rate of 0.05, 45 traits were returned as having some direction of causality with years of schooling (Figure 4b). Several of these putative relationships appear to be plausible, for example the influence of years of schooling on lower risk of Alzheimer's disease, which has recently been reported [@Anderson2017]. There is also a protective effect of more years of schooling on lower risk of myocardial infarction, which has been supported by detailed follow up analyses [] However, it is clear that there are limitations in relying on MR as a panacea for causal inference. For example, bi-directional causal relationships with college completion exist because of the similarity in trait definitions; and there are several traits (e.g. childhood intelligence and birth weight) which appear to be causally downstream of years of schooling despite this being temporally impossible. Though explanations could be conjured for this association, for example parents' schooling influences childhood intelligence, or the genetic instruments are shared across the two traits due to a third unmeasured trait, an intuitive interpretation of causality cannot be applied because the genetic instruments used proxy the liability to achieve some level of educational attainment, and not necessarily the experience of having attained a particular level of edication.

The facility to perform these scans has been automated in the MR of 'everything vs everything' (MR-EvE) web application, [http://eve.mrbase.org](http://eve.mrbase.org).

@@

The trend of increasing GWAS sample sizes continues, but while the opportunity that this affords MR to be furnished with more instruments is typically welcomed, here we have shown that it is invalid instruments, exhibiting horizontal pleiotropy or reverse causation, that are more likely to be identified than valid ones. Strategising how MR is to be used in practice, therefore, must consider horizontal pleiotropy to be the rule rather than the exception. Instrumental variables in MR are typically chosen blind, with GWAS significance being the only criterion. We illustrated how confounders and reverse causal associations can easily lead to invalid instrument selection in MR and that most randomly generated patterns of horizontal pleiotropy cannot be adequately accounted for by any single MR method.

>>>>

Mendelian randomization (MR) [@DaveySmith2003; @DaveySmithHemani2014] exploits vertical pleiotropy to infer the causal relationships between phenotypes. Suppose that one trait (the exposure) causally influences another (the outcome). If a SNP influences the outcome through the exposure then the SNP is exhibiting vertical pleiotropy (see Box 1). Such a genetic variant is considered to be a valid instrumental variable if it only influences the outcome through the exposure (the exclusion restriction assumption). Vertical pleiotropy can be exploited to mimic a randomised controlled trial, enabling a causal estimate to be made by comparing the outcome phenotypes between those individuals that have the exposure-increasing allele against those who do not, although not without caveats [@Holmes2017]. Multiple independent genetic variants for a particular exposure can be used jointly to improve causal inference, because a) each variant represents an independent natural experiment, and an overall causal estimate can be obtained by meta-analysing the single estimates from each instrument; and b) potential bias arising from the exclusion restriction assumption can be detected or corrected by evaluating the consistency of effects across instruments [@Bowden2015; @Bowden2016b; @Bowden2017; @Hartwig2017a; @Hemani2017; @Verbanck2017].

@@

Problems with obtaining unbiased causal effects can arise, however, if the genetic instruments exhibit horizontal pleiotropy, where they influence the outcome through a pathway other than the exposure (Box 1). The extent of this phenomenon is not to be understated, and many methods have been developed that attempt to reliably obtain unbiased causal estimates under specific models of horizontal pleiotropy [@Bowden2015; @Bowden2016b; @Bowden2017; @Hartwig2017a; @Verbanck2017; @Zhu2017]. It is considered best practice to triangulate estimates from a range of MR methods (and other experimental designs) when presenting causal estimates so as to assess the potential for different horizontal pleiotropy models and other violations of MR assumptions [@Lawlor2017]. However there are circumstances in which it is desirable to consolidate across many methods to obtain a single MR estimate. First, it could be of interest to screen links between thousands of traits for being potential causal associations, in which case a critical evaluation of each causal estimate of interest may not be possible or convenient. Second, though the simplest method, based on inverse-variance weighted (IVW) meta analysis [@Pierce2013], is most statistically powerful under no horizontal pleiotropy, it can have high false positive or low true positive rates in the presence of horizontal pleiotropy compared to other methods. Pleiotropy has been hypothesised to be universal [@Wright1968], though the degree and nature (e.g. horizontal versus vertical, see Box 1) to which this may be the case is contested [@Wagner2011; @Hill2012a], hence defaulting to the IVW method in the first instance and using other methods as sensitivity analyses may not be appropriate. Third, if different methods disagree it is not possible to know which is correct because the true nature of horizontal pleiotropy exhibited by the instruments is not known. Fourth, the available methods do not cover all possible models of horizontal pleiotropy, and therefore an automated method for instrument selection may be necessary.

@@

**Mode-based methods:** The mode-based estimator clusters the instruments into groups based on similarity of causal effects, and returns the final causal effect estimate based on the cluster that has the largest number of instruments [@Hartwig2017a]. There are four implementations of this method: the simple and the weighted mode, each weighted with or without the assumption of no measurement error in the exposure estimates (NOME). The simple mode is the unweighted mode of the empirical density function of causal estimates, whereas the weighted mode is weighted by the inverse variance of the outcome effect. The bandwidth parameter was set to 1 by default.

@@

We also performed a look up of associations that causally influence years of schooling [@Okbay2016], or that years of schooling itself causally influences. Using a false-discovery rate of 0.05, 45 traits were returned as having some direction of causality with years of schooling (Figure 4b). Several of these putative relationships appear to be plausible, for example the influence of years of schooling on lower risk of Alzheimer's disease, which has recently been reported [@Anderson2017]. There is also a protective effect of more years of schooling on lower risk of myocardial infarction, which has been supported by detailed follow up analyses [@Tillmann2017].

However, it is clear that there are limitations in relying on MR as a panacea for causal inference. For example, bi-directional causal relationships with college completion exist because of the similarity in trait definitions; and there are several traits (e.g. childhood intelligence and birth weight) which appear to be causally downstream of years of schooling despite this being temporally impossible. Though explanations could be conjured for this association, for example parents' schooling influences childhood intelligence, or the genetic instruments are shared across the two traits due to a third unmeasured trait, an intuitive interpretation of causality cannot be applied because the genetic instruments used proxy the liability to achieve some level of educational attainment, and not necessarily the experience of having attained a particular level of edication.

The facility to perform these look ups has been automated in the MR of 'everything vs everything' (MR-EvE) web application, [http://eve.mrbase.org](http://eve.mrbase.org).

@@

The trend of increasing GWAS sample sizes continues, but while the opportunity that this affords MR to be furnished with more instruments is typically welcomed, here we have demonstrated that it is invalid instruments, exhibiting horizontal pleiotropy or reverse causation, that may be more likely to be identified than valid ones. Strategising how MR is to be used in practice, therefore, must consider some presence of horizontal pleiotropy to be the rule rather than the exception. Instrumental variables in MR are typically chosen blind, with GWAS significance being the only criterion. We illustrated how confounders and reverse causal associations can easily lead to invalid instrument selection in MR and that most randomly generated patterns of horizontal pleiotropy cannot be adequately accounted for by any single MR method.
